---
title: "T. rex size distributions"
date: "2025-02-15"
categories: [dinosaurs, r]
---

I ought to be spending this weekend working on a risk intelligence LLM, but I've stumbled across something that brings my inner child immense joy and such things cannot be ignored. It is this 2019 paper ["An Older and Exceptionally Large Adult Specimen of Tyrannosaurus rex"](https://anatomypubs.onlinelibrary.wiley.com/doi/10.1002/ar.24118) by Persons et al. First it's cool to see paleontology in action, instead of filtered through talking head documentaries. Second, the diligent authors have personally measured 12 T. rex specimens, which is a decent chunk of the [29 discovered to-date](https://en.wikipedia.org/wiki/Specimens_of_Tyrannosaurus), and so we have a cool dataset to play with.

```{r}
#| warning: false
#| label: tbl-dataset
#| tbl-cap: T. rex bone measurements in millimetres (Table 1 from Persons et al 2019).
#| code-fold: true
library(tidyverse)
library(knitr)

t.rex <- read_tsv("t.rex-persons.tsv", col_types = "ccnnnnnnnnnnnnnnn") |>
  mutate(Morph = factor(Morph, levels=c("Robust", "Gracile")))
kable(t.rex, format.args = list(decimal.mark = ".", big.mark = ","))
```

The most interesting bone is the femur, being as it's more likely to be present and is used to estimate body mass. An interesting thing about T. rex is that there are two morphs, _gracile_ and _robust_, which I understand to be layman's terms for "elf-like" and "chonky bois".


```{r}
ggplot(t.rex) +
  aes(x = `Femur length`, y = `Femur circumference`, colour = Morph, label = Specimens) +
  geom_point() +
  geom_label(nudge_y = 10) +
  scale_x_continuous(limits=c(1000, 1400)) +
  scale_y_continuous(limits=c(400, 650)) +
  theme_minimal()
```