---
title: Gaussian Processes Again
subtitle: Yet another low-budget sequel
date: "2025-11-24"
categories: [gaussian, bayesian, stan, r]
freeze: false
code-fold: false
---

One nice aspect of blogging about the things you've just learned is that when, inevitably, you forget those things, reading back your old posts can bring you up to speed pretty quickly. It's like a slight groove has been worn and you can slip back into it more easily. (Now I realise why we say "get back in the groove" - it's nothing to do with funk after all! 🤦‍♂️)

Anyway, one such thing that I learned and forgot is Gaussian Processes. My [previous post](../gaussian-processes/index.qmd) helped, but I think that particular groove could be deeper. So here we go. This time let's do it backwards.

# OK so why Gaussian Processes?

Here's the problem you face: you have some noisy data generated by an unknown (continuous, real-valued, sexy) function. You can't abide the mystery and need to map out the unobserved regions of the function.

```{r}
#| code-fold: true
set.seed(42)
library(tidyverse)

```

```{r}
#| label: fig-data
#| fig-cap: Noisy observations (red points) of the mystery function (dashed line).
#| warning: false
mystery.function <- \(x) 1.1 * sin(x**0.5)

observations <- tibble(x = seq(0, 10, 0.5)) |>
  mutate(
    y = mystery.function(x) + rnorm(length(x), mean = 0, sd = 0.25),
  )

ggplot(observations) +
  aes(x = x, y = y) +
  geom_point(colour = "red", size = 3) +
  geom_function(fun = mystery.function, colour = "black", linetype = "dashed") +
  scale_y_continuous(limit = c(0, 1.5))
```

You could just fit a line, sure. If you have a good idea about the function that's producing your data you might as well do a regression to fit a line.

```{r}
#| label: fig-fit
#| fig-cap: Linear model fit (solid blue line) on the observations from the mystery function with 95% confidence interval (grey region).
ggplot(observations) +
  aes(x = x, y = y) +
  geom_point(colour = "red", size = 3) +
  geom_function(fun = mystery.function, colour = "black", linetype = "dashed") +
  geom_smooth(method = "lm", formula = y ~ sin(x**0.5)) +
  scale_y_continuous(limit = c(0, 1.5))
```

If we weren't being masters of this toy universe, able to see the coefficient of the true mystery function, we might be happy enough with that fit. But we are, and so we aren't. Instead we can see that the fit misses the maximum of the true function.

The confidence interval is wide there anyway. If we had a little more data in that region, we could improve the fit.

```{r}
#| label: fig-more-obs
#| fig-cap: The same model but with additional data points in [2.0, 3.0). The fit is improved.

observations.extra <- tibble(
  x = seq(2.0, 3.0, 0.1)
) |>
  mutate(
    y = mystery.function(x) + rnorm(length(x), 0, 0.25)
  )

ggplot(
  bind_rows(
    observations,
    observations.extra
  )
) +
  aes(x = x, y = y) +
  geom_point(colour = "red", size = 3) +
  geom_function(fun = mystery.function, colour = "black", linetype = "dashed") +
  geom_smooth(method = "lm", formula = y ~ sin(x**0.5)) +
  scale_y_continuous(limit = c(0, 1.5))
```

If we didn't have a good idea about the mystery function, we could fit anything else, like Loess.

```{r}
#| label: fig-fit
#| fig-cap: Loess fit (solid blue line) on the observations from the mystery function with 95% confidence interval (grey region).
ggplot(observations) +
  aes(x = x, y = y) +
  geom_point(colour = "red", size = 3) +
  geom_function(fun = mystery.function, colour = "black", linetype = "dashed") +
  geom_smooth() +
  scale_y_continuous(limit = c(0, 1.5))
```

# The Bayesian aproach

What we've done so far is the non-Bayesian approach to the problem. Why bother with the Bayesian approach (Gaussian Processes) at all?

It's just the usual reasons for choosing Bayesian models:

1. For a better understanding of the uncertainty.
2. To simplify probabilistic calculations (e.g. finding the probability that `f(2) > f(3)`).
3. It gives you the opportunity to introduce your prior knowledge about parameters.

Getting your head around the Bayesian approach is a little challenging though. We have to forget the idea of optimising coefficients of some linear formula to fit our model. Instead we are estimating possible locations of the line. We don't care about the formula any more.

Rather than a formula, we can use a Gaussian Process (GP). The GP is a distribution that generates "functions" (lines). How so? It's a multivariate normal distribution, that can generate a random value for many dimensions.

If you imagine that each dimension is a point on the x-axis, a sample from the multivariate normal distribution describe a random line, or some unknown function, to look at it the other way.

```{r}
# label: fig-random-draw
# fig-cap: Random draw from multivariate normal distribution, plotted as a line.
n.dims <- 16

# Sigma specifies the covariance matrix of the variables.
# This identify matrix means each dim is completely unrelated.
sigma <- matrix(0, nrow = n.dims, ncol = n.dims)
for (i in 1:n.dims) {
  sigma[i, i] = 1
}

random.draw <- MASS::mvrnorm(
  n = 1, # n.samples
  mu = rep(0, n.dims), # all centered on 0
  Sigma = sigma
)

ggplot(
  data.frame(x = 1:n.dims, y = random.draw)
) +
  aes(x = x, y = y) +
  geom_line()
```

The _kernel_ (sigma in the code above) controls the shape of the functions generated.

```{r}
#| label: fig-gp
#| fig-cap: Examples of functions drawn from a Gaussian Process with some fixed hyper parameters.

# Kernel
sq_exp_cov <- function(x, lambda, alpha) {
  n <- length(x)
  K <- matrix(0, n, n)
  for (i in 1:n) {
    for (j in 1:n) {
      diff <- sqrt(sum((x[i] - x[j])^2))
      K[i, j] <- alpha^2 * exp(-diff^2 / (2 * lambda^2))
    }
  }
  K
}

# TODO sample from GP for one alpha and lambda, plot it
```

