---
title: Gaussian Processes Again
subtitle: Yet another low-budget sequel
date: "2025-11-28"
categories: [gaussian, bayesian, stan, r]
freeze: false
code-fold: false
---

One nice aspect of blogging about the things you've just learned is that when, inevitably, you forget those things, reading back your old posts can bring you up to speed pretty quickly. It's like a slight groove has been worn and you can slip back into it more easily. (Now I realise why we say "get back in the groove" - it's nothing to do with funk after all! 🤦‍♂️)

Anyway, one such thing that I learned and forgot is Gaussian Processes. My [previous post](../gaussian-processes/index.qmd) helped, but I think that particular groove could be deeper. So here we go. This time let's do it backwards.

# OK so why Gaussian Processes?

Here's the problem you face: you have some noisy data generated by an unknown (continuous, real-valued, sexy) function. You can't abide the mystery and need to map out the unobserved regions of the function.

```{r}
#| code-fold: true
set.seed(42)
library(tidyverse)
```

R is our weapon of choice, as usual.

```{r}
#| label: fig-data
#| fig-cap: Noisy observations (red points) of the mystery function (dashed line).
#| warning: false
mystery.function <- \(x) 1.1 * sin(x**0.5)

observations <- tibble(x = seq(0, 10, 0.5)) |>
  mutate(
    y = mystery.function(x) + rnorm(length(x), mean = 0, sd = 0.25),
  )

ggplot(observations) +
  aes(x = x, y = y) +
  geom_point(colour = "red", size = 3) +
  geom_function(fun = mystery.function, colour = "black", linetype = "dashed") +
  scale_y_continuous(limit = c(0, 1.5))
```

You could just fit a line, sure. If you have a good idea about the function that's producing your data you might as well do a regression to fit a line.

```{r}
#| label: fig-fit
#| fig-cap: Linear model fit (solid blue line) on the observations from the mystery function with 95% confidence interval (grey region).
ggplot(observations) +
  aes(x = x, y = y) +
  geom_point(colour = "red", size = 3) +
  geom_function(fun = mystery.function, colour = "black", linetype = "dashed") +
  geom_smooth(method = "lm", formula = y ~ sin(x**0.5)) +
  scale_y_continuous(limit = c(0, 1.5))
```

If we weren't being masters of this toy universe, able to see the coefficient of the true mystery function, we might be happy enough with that fit. But we are, and so we aren't. Instead we can see that the fit misses the maximum of the true function.

The confidence interval is wide there anyway. If we had a little more data in that region, we could improve the fit.

```{r}
#| label: fig-more-obs
#| fig-cap: The same model but with additional data points in [2.0, 3.0). The fit is improved.

observations.extra <- tibble(
  x = seq(2.0, 3.0, 0.1)
) |>
  mutate(
    y = mystery.function(x) + rnorm(length(x), 0, 0.25)
  )

ggplot(
  bind_rows(
    observations,
    observations.extra
  )
) +
  aes(x = x, y = y) +
  geom_point(colour = "red", size = 3) +
  geom_function(fun = mystery.function, colour = "black", linetype = "dashed") +
  geom_smooth(method = "lm", formula = y ~ sin(x**0.5)) +
  scale_y_continuous(limit = c(0, 1.5))
```

If we didn't have a good idea about the mystery function, we could fit anything else, like Loess. But whatever we fit, we're making a lot of assumptions about the function and could end up quite far from the mark.

```{r}
#| label: fig-fit
#| fig-cap: Loess fit (solid blue line) on the observations from the mystery function with 95% confidence interval (grey region).
ggplot(observations) +
  aes(x = x, y = y) +
  geom_point(colour = "red", size = 3) +
  geom_function(fun = mystery.function, colour = "black", linetype = "dashed") +
  geom_smooth() +
  scale_y_continuous(limit = c(0, 1.5))
```

# The Bayesian aproach

What we've done so far is the non-Bayesian approach to the problem. Why bother with the Bayesian approach (Gaussian Processes) at all?

- First, because we don't necessarily have a good idea what our function is. It could be very nonlinear. A Gaussian Process is a non-parametric model, i.e. it doesn't assume any specific form of the function.

- Second, as usual with Bayesian models, you can get a richer understanding of uncertainty that you can with linear models and confidence intervals. It's also more suited to low-data scenarios.

Getting your head around the Bayesian approach can be a little challenging though. We have to forget the idea of optimising coefficients of some linear formula to fit our model. Instead we are estimating possible locations of the line. We don't care about the formula any more.

Rather than a formula, we can use a Gaussian Process (GP). The GP is a distribution that generates "functions" (lines). How so? It's a multivariate normal distribution, that can generate a random value for many dimensions.

If you imagine that each dimension is a point on the x-axis, a sample from the multivariate normal distribution describe a random line, or some unknown function, to look at it the other way.

```{r}
# label: fig-random-draw
# fig-cap: Random draw from multivariate normal distribution, plotted as a line.
n.dims <- 16

# Sigma specifies the covariance matrix of the variables.
# This identify matrix means each dim is completely unrelated.
sigma <- matrix(0, nrow = n.dims, ncol = n.dims)
for (i in 1:n.dims) {
  sigma[i, i] = 1
}

random.draw <- MASS::mvrnorm(
  n = 1, # n.samples
  mu = rep(0, n.dims), # all centered on 0
  Sigma = sigma
)

ggplot(
  data.frame(x = 1:n.dims, y = random.draw)
) +
  aes(x = x, y = y) +
  geom_line()
```

The _kernel_ (sigma in the code above) controls the shape of the functions generated. In the example above it's an identify matrix, but a more conventional choice would be squared exponential covariance. This generates a nice variety of functions. The two hyperparameters alpha and lambda control something like the amplitude and frequency of the generated functions.

```{r}
#| label: fig-gp
#| fig-cap: Examples of functions sampled from a Gaussian Process with alpha = 1 and lambda = 2.

# Kernel
sq_exp_cov <- function(x, lambda, alpha) {
  n <- length(x)
  K <- matrix(0, n, n)
  for (i in 1:n) {
    for (j in 1:n) {
      diff <- sqrt(sum((x[i] - x[j])^2))
      K[i, j] <- alpha^2 * exp(-diff^2 / (2 * lambda^2))
    }
  }
  K
}

plot_random_gp <- function(lambda, alpha) {
  n.samples <- 5
  x <- seq(0, 9, 0.1)
  samples <- MASS::mvrnorm(
    n = n.samples,
    mu = rep(0, length(x)),
    Sigma = sq_exp_cov(x, lambda = lambda, alpha = alpha)
  )

  t(samples) |>
    as_tibble() |>
    mutate(x = x) |>
    pivot_longer(-x, names_to = "sample", values_to = "y") |>
    ggplot() +
    aes(x = x, y = y, group = sample, colour = sample) +
    geom_line()
}
plot_random_gp(lambda = 2, alpha = 1)
```

So we have this mystery function generator! It can propose possible functions, and we can see how well they fit the data. We can also predict values for these functions at points that we don't have observations for.

Time for Stan. As per the last blog, this Stan code was taken from an excellent [course](https://carpentries-incubator.github.io/statistical-probabilistic-programming-r/instructor/gaussian-processes.html) on Carpentries.

I maintain that the best way to read Stan is back-to-front, so here is some invalid Stan that I've inverted and annotated.

```{stan}
#| label: stan
model {
  // f_x is sampled from the GP.
  // Remember it's a vector of f(x), not parameters describing f.
  f_x ~ multi_normal(rep_vector(0, n), K);

  // Likelihood is evaluated on f(x) plus normally distributed noise.
  y_obs ~ normal(f[1 : n_obs], sigma);

}
parameters {
  // Our outputs will be f(x) for all x (our observations, x_obs, and the predictions we want, x_pred)
  vector[n] f_x;
}
data {
  // Observed data
  int n_obs;
  array[n_obs] real x_obs;
  array[n_obs] real y_obs;
  
  // Observation error - pick a value
  real<lower=0> sigma;
  
  // x values for which we aim to predict f(x)
  int n_pred;
  array[n_pred] real x_pred;
  
  // Hyperparameters for the kernel
  real alpha;
  real lambda;
}
transformed data {
  // We join the x observations and desired x prediction points
  // because we want f(x) for both observed data and new predictions
  int n = n_obs + n_pred;
  
  array[n] real x;
  x[1 : n_obs] = x_obs;
  x[(n_obs + 1): n] = x_pred;
  
  // We calculate the Kernel values for all observed x
  matrix[n, n] K;
  K = gp_exp_quad_cov(x, alpha, lambda);
  
  // Add "nugget" on diagonal for numerical stability
  for (i in 1 : n) {
    K[i, i] = K[i, i] + 1e-6;
  }
}
```

The actual valid Stan file is [here](./gp.stan). An optimised version with Cholesky decomposition is [here](./gp-cholesky.stan), which is the one we'll run.

```{r}
#| label: stan
library(cmdstanr)

check_cmdstan_toolchain(fix = TRUE, quiet = TRUE)

x.pred <- seq(0, 10, by = 0.25)

model <- cmdstan_model(stan_file = "gp-cholesky.stan", exe = "gp.stan.bin")
sample <- model$sample(
  list(
    n_data = nrow(observations),
    x_data = as.array(observations$x),
    y_data = as.array(observations$y),
    # Sigma here is our guess about the noise level.
    # It was sigma=0.25 in the normal distribution we put on observations.
    sigma = 0.5,
    n_pred = length(x.pred),
    x_pred = x.pred,
    alpha = 1,
    lambda = 2
  ),
  parallel_chains = 4,
  max_treedepth = 20,
  show_messages = FALSE # disabled to avoid polluting the blog post, should be TRUE
)

sample
```

We wrangle the samples into (x, y) values and plot them alongside the original observations. It now looks a bit like a Christmas decoration.

```{r}
#| label: fig-samples
#| fig-cap: Samples from the GP model (green lines) alongside the original observations (red points), with the true mystery function (black dashed).

plot_draws <- function(sample) {
  mat <- sample$draws(format = "draws_matrix")
  x.vals <- c(observations$x, x.pred)
  draws <- as_tibble(mat) |>
    # Every row is a draw, which we number
    mutate(draw = 1:n()) |>
    # Since each column is an observation f(x) for x indices, we pivot
    pivot_longer(starts_with("f"), names_to = "x", values_to = "y") |>
    # And map the x index back to an x value
    mutate(
      idx = as.numeric(str_extract(x, "[0-9]+")),
      x = x.vals[idx],
      y = as.numeric(y)
    )

  ggplot(observations) +
    aes(x = x, y = y) +
    geom_line(
      data = draws,
      mapping = aes(group = draw),
      alpha = 0.01,
      colour = "darkgreen"
    ) +
    geom_point(colour = "red", size = 3) +
    geom_function(fun = mystery.function, colour = "black", linetype = "dashed")
}

plot_draws(sample)
```

The Bayesian model has approximated something sinusoidal. It fits the data pretty well, but has been tripped up by our noise.

What would have happened if we had different hyperparameters? Let's increase lambda so that our GP produces less wiggly functions. (It's Friday night, you're not getting a more technical description than that.)

```{r}
#| label: fig-random-gp-lambda-5
#| fig-cap: Example draws from a GP with the sq exp cov kernel and lambda = 5 and alpha = 1. They're much less "wiggly".
plot_random_gp(lambda = 5, alpha = 1)
```

Let's draw again with this calmer GP.

```{r}
#| label: fig-samples-2
#| fig-cap: Samples from the GP model (green lines) alongside the original observations (red points), with the true mystery function (black dashed). But this time we increased lambda and reduced the oscillations in our sampled functions. It gets closer to the true mystery function.
sample2 <- model$sample(
  list(
    n_data = nrow(observations),
    x_data = as.array(observations$x),
    y_data = as.array(observations$y),
    # Sigma here is our guess about the noise level.
    # It was sigma=0.25 in the normal distribution we put on observations.
    sigma = 0.5,
    n_pred = length(x.pred),
    x_pred = x.pred,
    alpha = 1,
    lambda = 5
  ),
  parallel_chains = 4,
  max_treedepth = 20,
  show_messages = FALSE # disabled to avoid polluting the blog post, should be TRUE
)

plot_draws(sample2)
```

That's generally closer to the mystery function! If we had some prior knowledge of the function - but not its formula - this would be useful.